Using Random Walk Simulations to Calculate Ground State Energies in Quantum Physics

### 2020-04-19 Paper/main.tex
Rewrote Results section

% *** CITATION PACKAGES ***
\usepackage[style=ieee]{biblatex} 
\bibliography{random_walk.bib}    %your file created using JabRef

@@

% *** PDF, URL AND HYPERLINK PACKAGES ***
\usepackage{url}
\hyphenation{op-tical net-works semi-conduc-tor}
\usepackage{graphicx}  %needed to include png, eps figures
\usepackage{float}  % used to fix location of images i.e.\begin{figure}[H]

@@

\section{Results and Discussion}
The simulation was first run for an infinite square well with boundary $J = 20$,
with 1000000 walkers. The step on which each walker died was recorded and then
the normalised histogram in Figure \ref{fig:hist} was constructed. The histogram
has been normalised, and is hence a probability distribution showing the
probability that a walker will die on a particular step. As we expect, the peak
is far from the origin, since for a walker to die in an inifinite square well
with $J=20$, they would have to walk at least 20 steps. It is also highly
unlikely that a walker would walk 20 steps in the same direction
consecutively. Thus, the peak is also at a number of steps greater than $J$. We
expect the shape of the decay after the peak to be exponential. We can see this
by plotting the histogram on a logarithmic scale.

Figure \ref{fig:loghist} presents the same histogram data with a logarithmic y
scale. In this figure, a straight line is evident in the data, showing the decay
was indeed exponential after the peak. The peak probability being away fro the
origin is also clearer in this figure. Figure \ref{fig:loghist} shows a
``widening'' of the straight line at higher step numbers. This is because there
is a greater variance in the number of walkers surviving to this number as it is
highly unlikely.

By taking the cumulative sum of the histograms in Figure \ref{fig:expplots}, we
can ensure the peak probability begins at the origin. Figure \ref{fig:cumplots}
present the cumulative sums of the histograms. They represent the probability of
surviving at least a particular number of steps. Hence, the peak probability of
1 is at the origin, since there is a probability of 1 that a walker will survive
the first step.

Figure \ref{fig:cumhist} presents the cumulative summed histogram, with a fitted
exponential curve of the form of Equation \ref{eq:firsteq}. The fit is very
good, with $R^2 = 0.9999998$, very close to 1. There is also negligible
numerical error in the fit, i.e. an error smaller than the quoted precision of
the fit parameters. This is because the simulation was run for 1000000 walkers,
providing a large sample size. Figure \ref{fig:cumloghist} presents the same
histogram with a logarithmic y-scale and a fitted straight line of the form of
Equation \ref{eq:straightline}. This fit has the same $R^2$ value, as it is
simply the same fit on a different y-scale, and is thus equally good.

We observe some fluctuation in the data in the higher steps. This is because
there are so few walkers surviving to this many steps that there is more
significant variance in their relative numbers. It is possible to remove this
variance by introducing a cutoff number of steps, i.e. a maximum number of steps
at which the walker will automatically die, regardless of whether they reached a
boundary or not. Thus, there will not be such high variance in the number of
walkers surviving to high number of steps. This is presented in Figure
\ref{fig:cum_line_cutoff_plot}. While this fit is observed to be better, with a
higher $R^2$ value, some systematic error has been introduced since there will now
be a disproportionately high number of walkers dying at the cutoff step. Thus,
the final $\lambda J^2$ value will be less accurate. On a faster computer, with
more efficient code, it is possible to forgo the step cutoff and allow the
simulation to run until all the walkers die. This may take much longer but will
ensure there will not be such a systematic error introduced.

We can use the gradient of the fitted line in Figure \ref{fig:cumloghist} as the
$\lambda$ value, and compare $\lambda J^2$ with the theoretical value. However,
to determine the reliability of the calculated value, it is necessary to repeat
the procedure many times and take the average $\lambda$ and the standard
deviation as the error.

It is also necessary to investigate how varying the boundary $J$ affects the
accuracy of the final answer. Figure \ref{fig:multi_line_plot} presents the
cumulative summed histograms with their fitted lines for many different boundary
$J$ values. No step cutoff was introduced. The calculated $\lambda J^2$ value is
shown in the legend. \todo{truncate numbers} As shown, the gradient changes, but
this is to be expected since the expected highest number of steps would increase
as the boundary is larger, since there is more ``space'' for the walker to move
before reaching a boundary. As shown in the legend, the calculated $\lambda J^2$
values get closer and closer and closer to the theoretical value as $J$
increases. This is due to the fact that as $J$ increases, there are more walkers
surviving to larger number of steps. \todo{make plot of lambda J against J}.
Thus, we can say that using a larger value of $J$ will result in a more
accurate answer but will also be more computationally intensive and take
longer. Thus, we make a tradeoff between speed and accuracy and choose $J=20$
for the final investigation.

Running the simulation 10 times and taking the average $\lambda J^2$ with error
being standard deviation, we find: $\lambda J^2 = 1.2339 \pm 0.0008$. We quote
the precision to the decimal place at which there is variation from the
theoretical result of $1.2337$. We notice also that the error in the calculated
answer is small (since there is not much variation between repeated simulations)
and the theoretical result is within the range of the calculated answer.

\todo{circular well}
\begin{figure*}%[ht!]
  \centering
  \subfloat[Histogram of probability of dying on particlar step]
  {\includegraphics[width=0.5\textwidth]{images/exp_plot.pdf}
  \label{fig:hist}}
  \centering
  \subfloat[Histogram of probability of dying on particular step in a logarithmic scale]
  {\includegraphics[width=0.5\textwidth]{images/line_plot.pdf}
    \label{fig:loghist}}
  \caption{Histograms showing the Probability of dying on a particular step in
    an infinite square well with $J = 20$. As can be seen, the peak is away from the
    origin, as is expected since it is not likely for a walker to die on a step $n$
    where $n < J$. The exponential decay after the peak is evident, implying a
    straight line shape if plotted on a logarithmic scale, which can be seen on
    the right hand side plot. In this logarithmic scale plot, the peak being away
    from the origin is clearer.}
  \label{fig:expplots}
\end{figure*}

\begin{figure*}%[ht!]
  \centering
  \subfloat[Histogram of probability of surviving to at least a particlar step]
  {\includegraphics[width=0.5\textwidth]{images/cum_exp_plot.pdf}
  \label{fig:cumhist}}
  \centering
  \subfloat[Histogram of probability of surviving to at least a particular step
  in a logarithmic scale]
  {\includegraphics[width=0.5\textwidth]{images/cum_line_plot.pdf}
  \label{fig:cumloghist}}
  \caption{Histogram showing the Probability of surviving to at least a
      particular step in an infinite square well with $J = 20$. This is a
      cumulative sum of the bins in Figure \ref{fig:hist}. As can be seen, the
      peak is at the origin, since it is certain that no walker will die on the
      first step. An exponential decay curve provides a good fit, with an $R^2$
      value of 0.9999998, very close to 1. Plotting on a logarithmic scale shows
      the straight line fit, the gradient of which is $\lambda$, giving $\lambda
      J^2 = 1.2334$, with a negligible error.}
  \label{fig:cumplots}
\end{figure*}

\todo{fix centering of cum plot}

\begin{figure}%[!ht]
  \begin{center}
    \includegraphics[width=0.45\textwidth]{images/cum_line_plot_cutoff.pdf}
    \caption{Histogram plotted on a logarithmic scale showing the Probability of
      surviving to at least a particular step in an infinite square well with $J
      = 20$, and also introducing a maximum number of steps at 2000, after which
      all walkers will die. This saves computational resources but introduces
      some systematic error, as there will now be a disproportionately high
      number of walkers in the 2000 steps bin.}
    \label{fig:cum_line_cutoff_plot}
  \end{center}
\end{figure}

\begin{figure}%[!ht]
  \begin{center}
    \includegraphics[width=0.45\textwidth]{images/multiplot.pdf}
    \caption{Histogram plotted on a logarithmic scale showing the Probability of
      surviving to at least a particualr step in different infinite square wells with
      different boundaries $J$. The $\lambda J^2$ values for each are shown. The
      values get closer to the theoretical value of $\frac{\pi^2}{8}$ as we increase
      the size of the well. This is likely due to more walkers surviving to a higher
      number of steps, as the well size is increased.}
    \label{fig:multi_line_plot}
  \end{center}
\end{figure}

@@

% if have a single appendix:
%\appendix[Proof of the Zonklar Equations]
% or
%\appendix  % for no appendix heading
% do not use \section anymore after \appendix, only \section*
% is possibly needed

% use appendices with more than one appendix
% then use \section to start each appendix
% you must declare a \section before using any
% \subsection or using \label (\appendices by itself
% starts a section numbered zero.)
%

\appendices
\section{Random walk Schr\"{o}dinger Equation Derivation}
\label{appendix:derivation}
Let us denote the probability that the walker is at position $j$ after $n$ steps
as $p(j,n)$ and the probabilty that the walker will die at position $j$ as
$a(j)$. We will denote the position-step number coordinates as $[j, n]$. Thus,
for a walker to have coordinates $[j, n+1]$, in the previous step, they must have
had coordinates $[j \pm 1, n]$. So we can say:

>>>>

% *** CITATION PACKAGES ***
\usepackage[style=ieee]{biblatex} 
\bibliography{random_walk.bib}

@@

% *** PDF, URL AND HYPERLINK PACKAGES ***
\usepackage{url}
\hyphenation{op-tical net-works semi-conduc-tor}
\usepackage{graphicx}  % needed to include png, eps figures
\usepackage{float}  % used to fix location of images i.e.\begin{figure}[H]

@@

\section{Results and Discussion}

\begin{figure*}%[ht!]
  \centering
  \subfloat[Histogram of probability of surviving to at least a particlar step]
  {\includegraphics[width=0.5\textwidth]{images/cum_exp_plot.pdf}
  \label{fig:cumhist}}
  \centering
  \subfloat[Histogram of probability of surviving to at least a particular step
  in a logarithmic scale]
  {\includegraphics[width=0.5\textwidth]{images/cum_line_plot.pdf}
  \label{fig:cumloghist}}
  \caption{Histogram showing the Probability of surviving to at least a
    particular step in an infinite square well with $J = 20$. The peak of 1 does
    not start at the origin but at step 100. This is because these steps were
    neglected in the plot as prior to this step the exponential shape is less
    clear, as explained by Figure \ref{fig:hist}. An exponential decay curve
    provides a good fit, with an $R^2$ value of 0.9999998, very close to
    1. Plotting on a logarithmic scale shows the straight line fit, the gradient
    of which is $\lambda$, giving $\lambda J^2 = 1.2334$, with a negligible
    error. The fluctuation in the higher step number regions is due to the large
    variance in the small number of walkers that die at these steps.}
  \label{fig:cumplots}
\end{figure*}

The simulation was first run for an infinite square well with boundary $J=20$
with 1000000 walkers. The number of walkers surviving at each step was recorded
and the histograms in Figure \ref{fig:cumplots} were created. The probability
that a walker will survive at a particular step is simply the 1 minus the
normalised cumulative sum of walkers dying at each step. The curve has been
fitted with an exponential function of the form in Equation
\ref{eq:firsteq}. This results in a very good fit, with an $R^2$ value of
0.9999998, indicating the data does not vary much from the curve of best
fit. This is likely due to the very large number of walkers used in this
experiment, as allowed by the performant C++ implementation.

We can see this better in Figure \ref{fig:cumloghist}, which is the same
exponential plot fitted on a logarithmic scale. The closeness of the data to the
line of best fit is evident here. The error in the fitted parameters is
negligible, as shown by the $R^2$ value. The gradient of the fitted line is
taken to be $\lambda$ and thus allows us to calculate $\lambda J^2 = 1.2334$,
compared to the theoretical value of $\frac{\pi^2}{8} = 1.2337$, quoting both
values to the precision at which they vary. Since the numerical error in the
fits is negligible, it cannot be propagated into an error in the final
answer. Thus, we get the error in the final answer from repeated runs of the
simulation and taking the standard deviation of the many values.

As we can observe in Figure \ref{fig:cumhist}, the peak of the curve does not
lie at the origin as we might expect from the fact that all walkers should
survive at least the first step. The reason for this is twofold:
\begin{itemize}
\item For a well with boundary $J$, it is not possible for the walker to die in
  a number of steps less than $J$
\item It is highly unlikely that the walker will die at a number of steps
  greater than but still close to $J$ since it would require walking in the same
  direction a dispropotionate number of times.
\end{itemize}

\begin{figure}%[!ht]
  \begin{center}
    \includegraphics[width=0.45\textwidth]{images/exp_plot.pdf}
    \caption{The probability of dying on a particular step in an infnite square
      well with $J=20$. As can be seen, the peak is away from the origin. This
      is because, for a well with boundary $J$, the minimum number of steps a
      walker must take is $J$. However, it is still unlikely that the walker
      will die on this minimum number of steps, as then the walker must walk in
      the same direction every time. Thus, the most probable step for the walker
      to die on is greater than $J$. It is for this reason that the exponential
      decay does not begin at the origin in Figure \ref{fig:cumhist}, meaning
      some earlier steps must be neglected in order to achieve a better fit.}
    \label{fig:hist}
  \end{center}
\end{figure}

These effects are evident in Figure \ref{fig:hist}, which presents the
probability that a walker will die at a particular number of steps. We can see
that the peak of the curve is very clearly at a step that is greater than 20,
which was the $J$ value used in this simulation. Thus, for low step numbers, the
shape of the curve is not exponential. For this reason we used a simple
algorithm to identify a suitable number of steps to skip before fitting the
exponential curve. It is for this reason that the histogram in Figure
\ref{fig:cumhist} does not start at the origin.

\begin{figure}%[!ht]
  \begin{center}
    \includegraphics[width=0.45\textwidth]{images/line_plot.pdf}
    \caption{The probability of dying on a particular step in an infnite square
      well with $J=20$, plotted on a logarithmic scale. As explained for Figure
      \ref{fig:hist}, the peak probability is on a step greater than $J$. This
      is clear in this plot. It is also evident that there is significantly more
      spread in the points at higher steps. This is due to there being
      significantly fewer walkers dying on these steps, so the relative
      differences are higher. This explains the fluctuations seen at higher
      steps in Figure \ref{fig:cumloghist}.}
    \label{fig:loghist}
  \end{center}
\end{figure}

We observe some fluctuation in the data in the higher steps in Figure
\ref{fig:cumloghist}. This is because there are so few walkers surviving to this
many steps that there is more significant variance in their relative numbers. We
can observe this in Figure \ref{fig:loghist}. At higher step numbers, the spread
is clearly larger due to the greater relative variance in the small number of
walkers surviving to this many steps. When this data is cumulatively summed to
calculate the probability of survival shown in Figure \ref{fig:cumloghist}, the
spread is present as the fluctuations.

\begin{figure*}%[ht!]
  \centering
  \subfloat[Probability of dying on a particlar step with a step cutoff]
  {\includegraphics[width=0.5\textwidth]{images/exp_plot_cutoff.pdf}
  \label{fig:expcutoff}}
  \centering
  \subfloat[Probability of surviving to at least a particular step in a
  logarithmic scale with a step cutoff]
  {\includegraphics[width=0.5\textwidth]{images/cum_line_plot_cutoff.pdf}
  \label{fig:cum_line_cutoff_plot}}
  \caption{Histograms showing the results of the simulation when a maximm number
  of steps is introduced. If a walker reaches this number of steps, they
  automatically die, regardless of position. This results in a
  disproportionately high number of walkers dying at this maximum number, 2000
  in this case, as shown by Figure \ref{fig:expcutoff}. However, this results in
  no fluctuation at higher number of steps, as shown in Figure
  \ref{fig:cum_line_cutoff_plot}, though some systematic error is introduced
  through this method.}
  \label{fig:cutoffplots}
\end{figure*}

It is possible to remove this variance by introducing a cutoff number of steps,
i.e. a maximum number of steps at which the walker will automatically die,
regardless of whether they reached a boundary or not. Thus, there will not be
such high variance in the number of walkers surviving to high number of
steps. This is presented in Figure \ref{fig:cum_line_cutoff_plot}. While this
fit is observed to be better, with a higher $R^2$ value, some systematic error
has been introduced since there will now be a disproportionately high number of
walkers dying at the cutoff step as can be seen in Figure
\ref{fig:expcutoff}. Thus, the final $\lambda J^2$ value will be less
accurate. On a faster computer, with more efficient code, it is possible to
forgo the step cutoff and allow the simulation to run until all the walkers
die. This may take much longer but will ensure there will not be such a
systematic error introduced. This is the method used in this investigation.

It is also necessary to investigate how varying the boundary $J$ affects the
accuracy of the final answer. Figure \ref{fig:multi_line_plot} presents the
cumulative summed histograms with their fitted lines for many different boundary
$J$ values. No step cutoff was introduced. The calculated $\lambda J^2$ value is
shown in the legend. As shown, the gradient changes, but this is to be expected
since the expected highest number of steps would increase as the boundary is
larger, since there is more space for the walker to move before reaching a
boundary. As shown in the legend, the calculated $\lambda J^2$ values get closer
and closer and closer to the theoretical value as $J$ increases. This is due to
the fact that as $J$ increases, there are more walkers surviving to larger
number of steps. Thus, we can say that using a larger value of $J$ will result
in a more accurate answer but will also be more computationally intensive and
take longer. Thus, we make a tradeoff between speed and accuracy and choose
$J=20$ for the final investigation.

Running the simulation 10 times and taking the average $\lambda J^2$ with error
being standard deviation, we find:
\begin{equation}
  \lambda J^2 = 1.2339 \pm 0.0008
  \nonumber
\end{equation}
We quote the precision to the decimal place at which there is variation from the
theoretical result of $1.2337$. We notice that the error in the calculated
answer is small (since there is not much variation between repeated simulations)
and the theoretical result is within the range of the calculated answer.

\todo{circular well}

\begin{figure}%[!ht]
  \begin{center}
    \includegraphics[width=0.45\textwidth]{images/multiplot1.pdf}
    \caption{Histogram plotted on a logarithmic scale showing the Probability of
      surviving to at least a particualr step in different infinite square wells
      with different boundaries $J$. The simulations were run with $10000$
      walkers The $\lambda J^2$ values for each are shown. The values get closer
      to the theoretical value of $\frac{\pi^2}{8} = 1.2337$ as we increase the
      size of the well. This is likely due to more walkers surviving to a higher
      number of steps, as the well size is increased.}
    \label{fig:multi_line_plot}
  \end{center}
\end{figure}

@@

\appendices
\section{Random walk Schr\"{o}dinger Equation Derivation}
\label{appendix:derivation}
Consider a one-dimensional lattice upon which a walker is undergoing a random
walker. Let us denote the probability that the walker is at position $j$ after
$n$ steps as $p(j,n)$ and the probabilty that the walker will die at position
$j$ as $a(j)$. We will denote the position-step number coordinates as $[j,
n]$. Thus, for a walker to have coordinates $[j, n+1]$, in the previous step,
they must have had coordinates $[j \pm 1, n]$. So we can say:
